---
title: "Case Study 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Divvy_Exercise_Full_Year_Analysis ###

# This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman (found here: https://artscience.blog/home/divvy-dataviz-case-study). The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: “In what ways do members and casual riders use Divvy bikes differently?”

# Adapted for case study 1 using Cyclistic2022 bike data

## We need these packages 

```{r}
library(tidyverse) # data wrangling
library(dplyr) # data wrangling
library(lubridate)  # work with dates
library(ggplot2)  # visualize data
library(scales) # visualize data

```



# Step 1: Collect Data

## Upload Divvy datasets (excel files) 

```{r}
dec2021 <- read.csv()
jan2022 <- read.csv()
feb2022 <- read.csv()
mar2022 <- read.csv()
apr2022 <- read.csv()
may2022 <- read.csv()
jun2022 <- read.csv()
jul2022 <- read.csv()
aug2022 <- read.csv()
sep2022 <- read.csv()
oct2022 <- read.csv()
nov2022 <- read.csv()
dec2022 <- read.csv()
```



# Step 2: Wrange Data and Place in a Single File

## Inspect the dataframes and look for incongruencies

```{r}
str(dec2021)
str(jan2022)
str(feb2022)
str(mar2022)
str(apr2022)
str(may2022)
str(jun2022)
str(jul2022)
str(aug2022)
str(sep2022)
str(oct2022)
str(nov2022)
str(dec2022)
```

## In Sep2022, end_station_id was numeric; changed data type to character for consistency

```{r}
sep2022 <- mutate(sep2022, end_station_id = as.character(end_station_id))
```

## Stack individual quarter's data frames into one big data frame

```{r}
all_trips <- bind_rows(dec2021, jan2022, feb2022, mar2022, apr2022, may2022, jun2022, jul2022, aug2022, sep2022, nov2022, dec2022)
```



# Step 3: Clean and Prepare Data

## Inspect new table

```{r}
colnames(all_trips)  # List of column names
nrow(all_trips)  # Number of rows are in data frame
dim(all_trips)  # Dimensions of the data frame
head(all_trips)  # See the first 6 rows of data frame
tail(all_trips) # See last 6 rows of data frame
str(all_trips)  # List of columns and data types 
summary(all_trips)  #Statistical summary of data
```

## There are a few problems we will need to fix:

### The data can only be aggregated at the ride-level. We will want to add some additional columns of data for day, month, and year to provide additional opportunities to aggregate the data.

### There are some rides where trip duration is negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We want to delete these rides.


## Add columns that list the date, month, day, and year of each ride
### Note that the final line changed day of week to a word instead of a number, from numeric to character

```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A") 
```

## Add a "ride_length" calculation to all_trips (in minutes)
### Changed data and hh:mm:ss format to seconds

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at, units = "mins")
```

## Convert "ride_length" from Factor to numeric and remove negative values from ride_length (indicates company quality check)

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
all_trips2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0),]
```



# Step 4: Descriptive Analysis

## Descriptive analysis on ride_length (all figures in minutes)

````{r}
mean(all_trips2$ride_length, na.rm = TRUE) #straight average (total ride length / rides)
median(all_trips2$ride_length, na.rm = TRUE) #midpoint number in the ascending array of ride lengths
max(all_trips2$ride_length, na.rm = TRUE) #longest ride
min(all_trips2$ride_length, na.rm = TRUE) #shortest ride
```

## You can condense the four lines above to one line using summary() on the specific attribute

```{r}
summary(all_trips2$ride_length)
```

## Compare members and casual users

```{r}
aggregate(all_trips2$ride_length ~ all_trips2$member_casual, FUN = mean)
aggregate(all_trips2$ride_length ~ all_trips2$member_casual, FUN = median)
aggregate(all_trips2$ride_length ~ all_trips2$member_casual, FUN = max)
aggregate(all_trips2$ride_length ~ all_trips2$member_casual, FUN = min)
```

## See the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips2$ride_length ~ all_trips2$member_casual + all_trips2$day_of_week, FUN = mean)
```

## Place days of the week in order

```{r}
all_trips2$day_of_week <- ordered(all_trips2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

## Average ride time by each day for members vs casual users

```{r}
aggregate(all_trips2$ride_length ~ all_trips2$member_casual + all_trips2$day_of_week, FUN = mean)
```

## Analyze ridership data by type and weekday

```{r}
all_trips2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by user type and weekday
  summarise(number_of_rides = n(),						#calculates the number of rides and average duration 
  average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```



# Visualize 

## Visualization by weekday

```{r}
all_trips2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  drop_na(weekday) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Weekday Usage by Membership Status", x = "Weekday", y = "Number of Rides") +
  scale_y_continuous(breaks = seq(0, 400000, 100000), labels = label_comma()) + 
  guides(fill = guide_legend(title="Membership")) +
  scale_fill_discrete(labels=c('Casual', 'Member')) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), axis.text = element_text(face = "bold")) 
```


## Visualization for average duration

```{r}
all_trips2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  drop_na(ride_length) %>%
  drop_na(weekday) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average Ride Duration by Membership Status", x = "Weekday", y = "Average Duration (minutes)") +
  scale_y_continuous(breaks = seq(0, 40, 5)) +
  guides(fill = guide_legend(title="Membership")) +
  scale_fill_discrete(labels=c('Casual', 'Member')) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), axis.text = element_text(face = "bold")) 
```

## Visualization for type of bike by membership

```{r}
all_trips2 %>% 
  drop_na(rideable_type) %>%
  filter(rideable_type == 'classic_bike' | rideable_type == 'electric_bike') %>%
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, rideable_type)  %>% 
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Bicycle Type by Membership Status", x = "Type of Bicycle", y = "Total Count",
       caption = "For clarity, casual riders account for 829,888 uses of \n classic bikes and for 806,673 uses of electric bikes.") +
  scale_x_discrete(labels = c("Classic", "Electric")) + 
  scale_y_continuous(breaks = seq(0, 1500000, 250000)) +
  guides(fill = guide_legend(title="Membership")) +
  scale_fill_discrete(labels=c('Casual', 'Member')) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), axis.text = element_text(face = "bold")) 
```



# Step 5: Summary file

```{r}
counts <- aggregate(all_trips2$ride_length ~ all_trips2$member_casual + all_trips2$day_of_week, FUN = mean)
write.csv(counts)
```
